---
title: "Heart Prediction Model"
author: "Farrah Johnson and Mariam Ikram"
date: "2026-04-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Intro
Women are underdiagnosed, often present atypical symptoms, medical datasets often underrepsent women...
Does dataset represent those patterns?
Heart disease is one of the leading causes of death worldwide, and the number 1 cause of death in the United States, about ... Americans from all ... (should i go into stats?) (citation) Understanding the its risk factors in pivotal for medical professionals and the general public ...

# Data Exploration: 
https://www.kaggle.com/datasets/fedesoriano/heart-failure-prediction/data
(what question we're answering and exploring, background on the data: where we got it from, the link, the author, the reasoning for the data being made available, important variables and their meanings)
I want to explore the difference in symptoms between males and females who have heart disease, because historically women have been underrepresented in medicine, their symptoms differ, and this contributes to misdiagnosis.

# Packages: 
tidyverse: data wrangling (dplyr), plotting (ggplot2) and general data manipulation 
ggplot2
dyplr
janitor
# Load Packages
```{r, message=FALSE, warning=FALSE, include=FALSE}

# tidyverse: data wrangling (dplyr), plotting (ggplot2), and general data manipulation (forcats)
library(tidyverse)

# janitor: cleans columns names and simple tabulations
library(janitor)

# knitr : Create nice tables in the report
library(knitr)

# kableExtra: Enhancing table appearance (border, alignment, etc.)
library(kableExtra)

#ggthemes: better themes for data distribution
library(ggthemes)

#ggdist: For visualizing data distribution in a more efficient way ????
library(ggdist)

```

```{r import}
# Import the CSV
heart_raw <- read.csv("heart.csv")

# Inspect raw structure
glimpse(heart_raw)
```

We first read in the CSV file and use glimpse() to inspect the raw imported data

# Data Cleaning
Now we will check and clean our data....

## Clean Column Names and Mutate dataset

```{r clean_up}
# ------------------
# Data Cleaning
# ------------------

# Clean column names
heart <- heart_raw %>% 
  clean_names() # converts column names to snake_case, lowercase, no spaces

# Recode Sex (F/M -> Female/Male)
heart <- heart %>% 
  mutate(sex = fct_recode(factor(sex),
                          "Female" = "F",
                          "Male" = "M"
                          ))

# Recode chest pain type abbreviations
heart <- heart %>% 
  mutate(chest_pain_type = fct_recode(factor(chest_pain_type),
                                      "Atypical Angina" = "ATA", 
                                      "Asymptomatic" = "ASY",
                                      "Non-Anginal Pain" = "NAP", 
                                      "Typical Angina" = "TA"
                                      ))


# Recode resting ECG type abbreviations
heart <- heart %>% 
  mutate(resting_ecg = fct_recode(factor(resting_ecg),
                                  "ST-T Abnormality" = "ST",
                                  "LV Hypertrophy" = "LVH",
                                  "Normal" = "Normal"
                                  ))

# Recode binary variables (0/1 or N/Y -> No/Yes)
heart <- heart %>% 
  mutate(across(c(fasting_bs, exercise_angina, heart_disease), ~ {
    # Turn the current column (.x) into a factor
    fct_col <- factor(.x)
    
    # Recode that factor
    fct_recode(fct_col,
               "Yes" = "1", "Yes" = "Y",
               "No" = "0", "No" = "N"
               )
  }))

# Check structure after cleaning
glimpse(heart)
```
 
We use `clean_names()` to standardize the column names. This makes it easier to reference variables in the rest of the analysis (for example, `chest_pain_type` instead of `ChestPainType`). 
Next we mutate our dataset, recoding clinical abbreviations with their full descriptive names. Now values such as `M` and `F` are now `Male` and `Female`, along with several others.
We use `fct_recode()` as opposed to simply converting variables to factors, because `fct_recode()` allows us to explicitly map each original value to a new, more readable label, improving clarity, reducing the chance of errors, and making the dataset easier to interpret in later summaries, analysis, and visualizations.

## Check for Missing Values

```{r missing_values}

# count missing values in each column
missing_vals <- colSums(is.na(heart))
missing_vals

```

After cleaning the data we check for missing values. Missing values indicate incomplete data which can skew summary statistics if not handled correctly. The results show this dataset has no missing values, therefore the data is complete requiring no further action in this regard.

## Check for Duplicates

```{r duplicate_check}

# Check for duplicates
duplicates <- heart %>% duplicated()

duplicates_table <- duplicates %>% table()

duplicates_table

```

Here we check for duplicates in the rows of the data set using duplicated(). Duplicates can occur when data is collected from multiple sources or from errors in entry, if duplicates were found I would remove them to ensure the integrity of the data. Since no duplicates were found, no further action is required. 

## Zero Value Check

```{r zero_value_check}

# Check how many cholesterol = 0
heart %>% summarize(zero_chol_vals = sum(cholesterol == 0, na.rm = TRUE))

# Replace those 0 cholesterol values with NA
heart <- heart %>% 
  mutate(cholesterol = na_if(cholesterol, 0))

# Check to see if 0 cholesterol values were replaced with NA 
heart %>% summarize(remaining_zeros = sum(cholesterol == 0, na.rm = TRUE))

#Check to see if resting_bp has any 0 values
heart %>% summarize(zero_resting_bp_vals = sum(resting_bp == 0, na.rm = TRUE))

# Replace those 0 resting_bp values with NA
heart <- heart %>% 
  mutate(resting_bp = na_if(resting_bp, 0))

# Check to see if 0 cholesterol and resting_bp values were replaced
heart %>% 
  summarize(
    zero_check_chol = sum(cholesterol == 0, na.rm = TRUE),
    zero_check_rest_bp = sum(resting_bp == 0, na.rm = TRUE))

```

While analyzing the dataset through boxplots and charts we noticed the dataset had variables that equaled to 0 in variables where a 0 is a physiological impossibility such as cholesterol, the data had 179 cholesterol values equal to 0, and resting blood pressure, where one of the values were a 0.
It can be assumed that those 0 values represent missing patient data, so we replaced all those 0 values with NA to accurately showcase the missing data in our dataset and analysis. 


#keep this part?
```{r, include = FALSE}

# takes you to the full database without including it in the knit

view(heart)
```


## Variable Dictionary

```{r variable_dictionary}
#-----------------------
# Variable Dictionary
#-----------------------

variable_dictionary <- tibble(
  variable = c(
    "age", "sex", "chest_pain_type", "resting_bp", "cholesterol", "fasting_bs",
    "resting_ecg", "max_hr", "exercise_angina", "oldpeak", "st_slope", "heart_disease"),
  description = c(
    "Age of patients in years",
    "Biological sex (Male/Female)",
    "Type of chest pain experienced (Typical Angina, Atypical Angina, Non-Anginal Pain, Asymptomatic)",
    "Resting blood pressure (mm Hg)",
    "Serum cholestrol level (mg/dL)",
    "Fasting blood sugar > 120 mg/dl (Yes/No)",
    "Resting electrocardiogram results (Normal, ST-T Abnormality, LV Hypertrophy)",
    "Maximum heart rate achieved during exercise",
    "Excercise-induced angina (Yes/No)",
    "ST depression induced by exercise relative to rest",
    "Slope of the peak ST segment",
    "Presence of heart disease (Yes/No)"
    )
  )

variable_dictionary %>% 
  kable(caption = "Variable Dictionary for Heart Dataset") %>% 
  kable_styling(bootstrap_options = c("striped", "bordered"))


```

** Clinical Background for Key Heart Variables

Several variables in this dataset come from electrocardiograms (ECG) (test that record the elecrical signals of the heart) and exercise stress tests. These exercises help determine how the heart responds to physical activity and whether blood flow to the heart is reduced. 

- ECG:
 
ST in the dataset refers to the ST segment (ST doesn't stand for anything, it is referring to the S and T points on the waves in the ECG.
- ST Segment/ ST-T Wave: The ST segment is part of the ECG waveform. Abnormal changes in this segment can indicate reduced blood flow to the heart (ischemia) or other electrical disturbances. 

- LV Hypertrophy (Left Ventricular Hypertrophy): occurs ventricle- the main pumping chamber of the heart- becomes thickened. Often resulting from long standing high blood pressure or other conditions that forces the heart to work harder. LV Hypertrophy is important because it increases the risk heart failure, and other cardiac events. 

- Oldpeak: This measure how much the ST segment drops during the exercise compared to rest. Higher values often indicate that the heart is not receiving enough oxygen.

- ST Slope: This describes the direction of the ST segment during peak exercise (upsloping, flat, downsloping). A flat or downsloping line is commonly associated with heart disease.

- Exercise Angina: Indicates whether the patient experienced chest pain during exercise, which can be a sign of reduced oxygen supply to the heart. 
-- Types of Anginal pain...

# Final Clean Dataset (condensed view)
```{r cleaned_condensed_dataset}
#-------------------------
# Sample of Cleaned Data
#-------------------------

heart %>% 
  slice_head(n = 10) %>% 
  kable(caption = "Sample of Cleaned Heart Dataset") %>% 
  kable_styling(bootstrap_options = c("striped", "bordered"), full_width = FALSE)

```

## Check for Outliers

To identify potential outliers, we created boxplots to evaluate the main numeric variables associated with heart disease in this dataset, including age, maximum heart rate, oldpeak, and resting blood pressure. Categorical variables such as chest pain type, resting ECG, and ST slope were excluded because they cannot have statistical outliers; instead, their distributions are examined later in the EDA section using bar charts.

### Age Outliers

```{r age_outliers}

heart %>%
  ggplot(aes(y = age)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Outliers in Age", y = "Age") +
  theme_minimal()

```

Age does not show any extreme outliers, meaning all the patients fall around a consistent age range.

### Resting Blood Pressure Outliers

```{r resting_bp_outliers}

heart %>%
  ggplot(aes(y = resting_bp)) +
  geom_boxplot(fill = "lightblue", na.rm = TRUE) +
  labs(title = "Outliers in Resting Blood Pressure", y = "Resting Blood Pressure") +
  theme_minimal()

```

Resting blood pressure shows several outliers above 150, and one unusually low value below a 100. A subset of patient differ significantly from the rest in their resting blood pressure reading. 

Note: We used "na.rm = TRUE" in geom_boxplot() to suppress the warning indicating a row was removed due to NA values created during data cleaning. Since the warning doesn't affect the visualization and is not indicative of anything significant it was silenced for a cleaner report.

### Max Heart Rate Outliers

```{r max_hr_outliers}

heart %>%
  ggplot(aes(y = max_hr)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Outliers in Max Heart Rate", y = "Max Heart Rate") +
  theme_minimal()

```

Max heart heart does have some extreme outliers in the lower end of the spectrum with heart rates below a 100, showing a small subset of patients had significantly reduced exercise capacity in comparison to the rest of the sample.  

### Oldpeak Outliers

```{r oldpeak_outliers}

heart %>%
  ggplot(aes(y = oldpeak)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Outliers in Oldpeak", y = "Oldpeak") +
  theme_minimal()

```

Oldpeak showed several extreme outliers above 2.5 and one unusually low value below -2.5. This indicates that a subset of patients experienced significantly greater and unusually negative ST segment depression during exercise. 

While we identified several outliers across the variables we will be using to analyze this dataset, we chose not to remove them. In many analyses outliers are removed to ensure the most accurate representation of the data. However, in this case, the outliers represent real patients with clinically significant conditions, removing them would eliminate important clinical variations leading to possibly incorrect and inaccurate representation and understanding of the population. Keeping them in ensures the analysis reflects the full range of the patients present with their unique variations.  

# EDA

We will now explore our dataset through visualizations and summary statistics.

### Count Bar Chart (Male vs Female)

```{r sex_count}
#-----------------------------------
# Count Bar Chart (Male vs Female)
#-----------------------------------
heart %>% 
  ggplot(aes(x = sex)) +
  geom_bar(fill = "darkgray") +
  labs(title = "Count of Male vs Female Patients",
       x = "Sex",
       y = "Count") +
  theme_minimal()

```

The bar chart shows the count of male vs female patients. This dataset contains significantly more males than females patients, creating a clear imbalance in the sample. This imbalance is important showing that raw counts alone can be misleading when comparing groups. It also hints at sampling bias or oversampling of male patients, which is common in meany heart data sets. Because of this imbalance proportion based visualizations are necessary later in the analysis to make fair comparisons between both genders. 

### Heart Disease Proportion Bar Chart (Male vs Female)

```{r heart_disease_by_sex}
#--------------------------------------------
# Heart Disease Proportion (Male vs Female)
#--------------------------------------------
heart %>% 
  ggplot(aes(x = sex, fill = heart_disease)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportion of Heart Disease by Sex",
       x = "Sex",
       y = "Percentage") +
  theme_minimal()

```

This proportion bar chart shows the prevalence of heart disease in males vs females. I used a proportion chart due to the raw count imbalance- a proportion bar chart is best to showcase a direct comparison, normalizing each group to a 100%, regardless of actual count. 
The proportion chart shows that about 63%  of males have heart disease in comparison to only 26% of females that do. This correlates with real world known heart disease patterns.

## Disease vs. No Disease

What symptoms are typically associated with heart disease? What symptoms distinguish heart disease from no heart disease?

### Chest Pain Type Distribution (Disease vs. No Disease)

```{r chest_pain_type_distribution}

heart %>% 
  ggplot(aes(x = chest_pain_type, fill = heart_disease)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Chest Pain Type by Heart Disease Status",
       x = "Chest Pain Type",
       y = "Percentage",
       fill = "Heart Disease") +
  theme_minimal()

```

This chart shows chest pain types between within each heart disease group. Among individual's with heart disease asymptomatic chest pain type is most prevalent, followed by typical angina, non-anginal pain, and atypical angina. This highlights that many patients with heart disease do not experience typical chest pain symptoms, marking it as a meaningful symptom to explore by gender.

### Oldpeak Distribution (Disease vs. No Disease)

```{r oldpeak_distribution}

heart %>% 
  ggplot(aes(x = heart_disease, y = oldpeak, fill = heart_disease)) +
  geom_boxplot() +
  labs(title = "Oldpeak by Heart Disease Status",
       x = "Heart Disease",
       y = "Oldpeak") +
  theme_minimal()

```

Individuals with heart disease tend to have a higher oldpeak in comparison to people with no heart disease, reflective of typical symptoms due to more pronounced ST‑segment depression during exercise. This pattern aligns with typical symptoms of ischemia and reinforces oldpeak as an important variable for distinguishing between the two groups.

### Max heart rate (Disease vs. No Disease)

```{r max_hr}

heart %>% ggplot(aes(x = heart_disease, y = max_hr, fill = heart_disease)) +
  geom_boxplot() +
  labs(title = "Max Heart Rate Distribution by Heart Disease Status",
       x = "Heart Disease Status", y = "Max Heart Rate") +
  theme_minimal()
 
```

Individuals with heart disease tend to have lower maximum heart rates than those without the condition, shown by the distribution of the box plot, in alignment with reduced cardiac functions. We can conclude that lower heart rates is a significant indicator of heart disease, but not the only indicator, as the graphs show an overlap.

### ST Slope (Disease vs. No Disease)

```{r ST_slope}

heart %>% 
  ggplot(aes(x = st_slope, fill = heart_disease)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "ST Slope by Heart Disease Status",
       x = "ST Slope", y = "Percentage", fill = "Heart Disease") +
  theme_minimal()

```

A down and flat ST slope is associated with patients with heart disease, this chart showcases patients with heart disease are prone to a down or flat ST slope with minimal variance, reinforcing its diagnostic relevance. 

### Resting ECG (Disease vs. No Disease)

```{r resting_ECG}

heart %>% 
  ggplot(aes(x = resting_ecg, fill = heart_disease)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Resting ECG Results by Heart Disease Status",
    x = "Resting ECG Result",
    y = "Percentage",
    fill = "Heart Disease"
  ) +
  theme_minimal()

```
Among people with heart disease, ST–T abnormalities are the most common Resting ECG finding, followed by left ventricular hypertrophy (LVH) and then normal ECG results. This indicates that many patients with heart disease show electrical evidence of ischemia or structural changes at rest. In contrast, individuals without heart disease show a higher proportion of normal ECGs, marking it as a clinically relevant marker between disease and no disease population

## Male vs Female (Disease vs No Disease)
Do these symptoms differ between men and women, accounting for disease vs. no disease.

### Resting ECG Results by Sex and Heart Disease
```{r }

heart %>% 
  ggplot(aes(x = resting_ecg, fill = heart_disease)) +
  geom_bar(position = "fill") +
  facet_wrap(~ sex) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Resting ECG Results by Sex and Heart Disease",
    x = "Resting ECG Result",
    y = "Percentage",
    fill = "Heart Disease") +
  theme_minimal()

```

Women and men present vastly different resting ECG pattern across disease groups. In women, the no‑disease group makes up the majority of the bar across all Resting ECG categories (ST–T abnormality, LV hypertrophy, and normal). The disease group is very small, so its portion of each bar is minimal. This means that even though women show ECG changes, these changes occur mostly in women without heart disease, reinforcing that Resting ECG abnormalities are not strong indicators of heart disease in women, reflecting clinical studies showing women tend to present with subtler or “nonspecific” ECG changes, making heart disease harder. 

In contrast, men show the expected clinical pattern: the disease group dominates, especially in ST–T abnormalities, followed by LV hypertrophy and then normal results- a pattern more consistent with typical heart disease presentation.


### Chest Pain Type by Sex and Heart Disease

```{r chest_pain_type_sex_disease}

heart %>% 
  ggplot(aes(x = chest_pain_type, fill = heart_disease)) +
  geom_bar(position = "fill") +
  facet_wrap(~ sex) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Chest Pain Type by Sex and Heart Disease",
       x = "Chest Pain Type",
       y = "Percentage",
       fill = "Heart Disease") +
  theme_minimal() +
  # angle text (prevent overlapping)
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

```
Asymptomatic is the most common chest pain type in the heart disease population, in both women and men.  However, the patterns for the remaining chest pain types differ greatly by sex.

In women, the other chest pain categories' (atypical angina, non‑anginal pain, and typical angina) distribution is relatively flat, showing little variation across chest pain types. Each make up less than 15% of cases, with atypical angina being the least common. This aligns with clinical evidence showing that women often present with non‑classic chest pain, which contributes to under‑recognition of heart disease in female patients

In men, the pattern is more distinct: after asymptomatic pain, typical angina is the next most common, followed by non‑anginal pain, with atypical angina being the least frequent, creating a clear gradient of chest pain types across the males with heart disease population, representing typically associated heart disease ischemic patterns and symptoms.


### Oldpeak by Sex and Heart Disease

```{r oldpeak_distribution_sex_disease}

heart %>% 
  ggplot(aes(x = heart_disease, y = oldpeak, fill = heart_disease)) +
  geom_boxplot() +
  facet_wrap(~ sex) +
  labs(title = "Oldpeak by Sex and Heart Disease",
       x = "Heart Disease",
       y = "Oldpeak",
       fill = "Heart Disease") +
  theme_minimal()

```

Women with no heart disease have a higher oldpeak than men with no heart disease, showing a greater ST segment depression during exercise in the non-disease group. However in the heart diseased population, women and men both have a similar old peak distribution. This aligns with clinical trials showing that old peak is generally an unreliable indicator of heart disease in women than in men.

### Max Heart Rate by Sex and Heart Disease

```{r max_hr_sex_disease}

heart %>% 
  ggplot(aes(x = heart_disease, y = max_hr, fill = heart_disease)) +
  geom_boxplot() +
  facet_wrap(~ sex) +
  labs(title = "Max Heart Rate by Sex and Heart Disease",
       x = "Heart Disease",
       y = "Max Heart Rate", 
       fill = "Heart Disease") +
  theme_minimal()

```
Max heart rate between men and women across the non-disease population is fairly similar. 

The heart disease population however shows a marked gender difference. Women with heart disease reach a higher max heart rate than men with heart disease. With the upper range values reaching over 150 while the men do not reach that level; aligning with evidence that women present different ischemic pattern in comparison to men.  

### ST Slope by Sex and Heart Disease

```{r ST_slope_sex_disease}

heart %>% 
  ggplot(aes(x = st_slope, fill = heart_disease)) +
  geom_bar(position = "fill") +
  facet_wrap(~ sex) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "ST Slope by Sex and Heart Disease)",
       x = "ST Slope", y = "Percentage", fill = "Heart Disease") +
  theme_minimal()

```

In this dataset, men with heart disease most often show a flat ST slope, followed by down, with up being least common. Women show the opposite pattern; downsloping is most common, then flat, and upsloping is rare.

Clinically, ST slope is known to be a stronger predictor of heart disease in women than oldpeak. This is reflective in the chart- over 15% of men with heart disease still show an upward slope (a pattern usually seen in healthier hearts), while fewer than 10% of women do, reinforcing the diagnostic importance of ST slope in identifying heart disease in women.

### Resting Blood Pressure by Sex and Heart Disease

```{r resting_bp_sex_disease}

heart %>% 
  ggplot(aes(x = heart_disease, y = resting_bp, fill = heart_disease)) +
  geom_boxplot() +
  facet_wrap(~ sex) +
  labs(title = "Resting Blood Presssure by Sex and Heart Disease",
       x = "Heart Disease",
       y = "Resting Blood Pressure", 
       fill = "Heart Disease") +
  theme_minimal()
ggplot(heart, aes(x = sex, y = age, fill = sex)) +
  geom_boxplot() +
  theme_minimal()
  
```
Women and men show similar resting blood pressure in the non‑disease group.
However, among individuals with heart disease, women present with higher resting blood pressure than men. This aligns with clinical evidence showing that elevated blood pressure is a stronger predictor of heart disease in women, with risk increasing even at levels as low as 110 mmHg, below the traditional 120 mmHg threshold. As women age and transition through menopause, their resting blood pressure often catches up to and surpasses men’s, sharply increasing cardiovascular risk.

# Summary 

Clinically, this aligns with evidence that women often present different ischemic patterns during exercise testing. Women may show fewer classic exercise‑induced heart‑rate limitations, which can make heart disease harder to detect using traditional stress‑test thresholds that were historically validated in male populations.

While this dataset does reflect real‑world gender differences — such as women having less typical chest pain and subtler ECG changes — it’s important to acknowledge the limitations.

The dataset is heavily skewed toward men, with nearly 80% male participants.

What’s striking is that this dataset was compiled from five different sources, and even after combining them, we still see a major gender imbalance.

This reinforces the gender inequality we talked about at the beginning: women are consistently underrepresented in cardiovascular research.

Because of this imbalance, the patterns we observed are meaningful, but they cannot be considered fully conclusive.

Instead, they highlight the need for more balanced, inclusive data in future studies..


























