---
title: "Heart Prediction Model"
author: "Farrah Johnson and Mariam Ikram"
date: "2026-04-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Intro: 
Women are underdiagnosed, often present a typical symptoms, medical datasets often underrepsent women...
Does dataset represent those patterns?
Heart disease is one of the leading causes of death worldwide, and the number 1 cause of death in the United States, about ... Americans from all ... (should i go into stats?) (citation) Understanding the its risk factors in pivotal for medical professionals and the general public ...

Data Exploration: (what question we're answering and exploring, background on the data: where we got it from, the link, the author, the reasoning for the data being made available, important variables and their meanings)
I want to explore the difference in symptoms between males and females who have heart disease, because historically women have been underrepresented in medicine, their symptoms differ, and this contributes to misdiagnosis.

Packages: 
tidyverse: data wrangling (dplyr), plotting (ggplot2) and general data manipulation 
ggplot2
dyplr
janitor #cleans columns names and simple tabulations

```{r, message=FALSE, warning=FALSE, include=FALSE}

# tidyverse: data wrangling (dplyr), plotting (ggplot2), and general data manipulation (forcats)
library(tidyverse)

# janitor: cleans columns names and simple tabulations
library(janitor)

# knitr : Create nice tables in the report
library(knitr)

# kableExtra: Enhancing table appearance (border, alignment, etc.)
library(kableExtra)

```

```{r import}
# Import the CSV
heart_raw <- read.csv("heart.csv")

# Inspect raw structure
glimpse(heart_raw)
```

We first read in the CSV file and use glimpse() to inspect the raw imported data

```{r clean_up}
# ------------------
# Data Cleaning
# ------------------

# Clean column names
heart <- heart_raw %>% 
  clean_names() # converts column names to snake_case, lowercase, no spaces

# Recode Sex (F/M -> Female/Male)
heart <- heart %>% 
  mutate(sex = fct_recode(factor(sex),
                          "Female" = "F",
                          "Male" = "M"
                          ))

# Recode chest pain type abbreviations
heart <- heart %>% 
  mutate(chest_pain_type = fct_recode(factor(chest_pain_type),
                                      "Atypical Angina" = "ATA", 
                                      "Asymptomatic" = "ASY",
                                      "Non-Anginal Pain" = "NAP", 
                                      "Typical Angina" = "TA"
                                      ))


# Recode resting ECG type abbreviations
heart <- heart %>% 
  mutate(resting_ecg = fct_recode(factor(resting_ecg),
                                  "ST-T Abnormality" = "ST",
                                  "LV Hypertrophy" = "LVH",
                                  "Normal" = "Normal"
                                  ))

# Recode binary variables (0/1 or N/Y -> No/Yes)
heart <- heart %>% 
  mutate(across(c(fasting_bs, exercise_angina, heart_disease), ~ {
    # Turn the current column (.x) into a factor
    fct_col <- factor(.x)
    
    # Recode that factor
    fct_recode(fct_col,
               "Yes" = "1", "Yes" = "Y",
               "No" = "0", "No" = "N"
               )
  }))

# Check structure after cleaning
glimpse(heart)
```
 
We use `clean_names()` to standardize the column names. This makes it easier to reference variables in the rest of the analysis (for example, `chest_pain_type` instead of `ChestPainType`). 
Next we mutate our dataset, recoding clinical abbreviations with their full descriptive names. Now values such as `M` and `F` are now `Male` and `Female`, along with several others.
We use `fct_recode()` as opposed to simply converting variables to factors, because `fct_recode()` allows us to explicitly map each original value to a new, more readable label, improving clarity, reducing the chance of errors, and making the dataset easier to interpret in later summaries, analysis, and visualizations.

```{r missing_values}

# count missing values in each column
missing_vals <- colSums(is.na(heart))
missing_vals

```
After cleaning the data we check for missing values. Missing values indicate incomplete data which can skew summary statistics if not handled correctly. The results show this dataset has no missing values, therefore the data is complete requiring no further action in this regard.

```{r duplicate_check}

# Check for duplicates
duplicates <- heart %>% duplicated()

duplicates_table <- duplicates %>% table()

duplicates_table

```
Here we check for duplicates in the rows of the data set using duplicated(). Duplicates can occur when data is collected from multiple sources or from errors in entry, if duplicates were found I would remove them to ensure the integrity of the data. Since no duplicates were found, no further action is required. 

```{r, include = FALSE}

# takes you to the full database without including it in the knit

view(heart)
```

```{r}
names(heart)
range(heart$age)
```




```{r clean_comparisons}

# Pulls out necessary data and changes the heart disease to a string to make it more decipherable 
# The 0 and 1 values in heart_disease instead of yes and no was driving me crazy. I cleaned it up and stored it in a variable.

heart_clean <- heart %>%
  mutate(heart_disease = factor(heart_disease,
                               levels = c(0, 1),
                               labels = c("No", "Yes")))

set.seed(123)
random_sample <- heart_clean %>%
  sample_n(20)
#I set a seed here so the data we're pulling is more random, as head(20) will only pull the first 20. 


random_sample %>%
  select(age, sex, heart_disease, exercise_angina, st_slope , oldpeak) %>%

  arrange(age) %>%
  head(20)

# gives an average for necessary data for a better comparison


heart_clean %>%
  select(age, sex, heart_disease, exercise_angina, st_slope , oldpeak) %>%
  
  group_by(heart_disease) %>%
  summarise("Average age" = mean(age),
            "Old Peak" = mean(oldpeak))


```

```{r variable_dictionary}
#-----------------------
# Variable Dictionary
#-----------------------

variable_dictionary <- tibble(
  variable = c(
    "age", "sex", "chest_pain_type", "resting_bp", "cholesterol", "fasting_bs",
    "resting_ecg", "max_hr", "exercise_angina", "oldpeak", "st_slope", "heart_disease"),
  description = c(
    "Age of patients in years",
    "Biological sex (Male/Female)",
    "Type of chest pain experienced",
    "Resting blood pressure (mm Hg)",
    "Serum cholestrol level (mg/dL)",
    "Fasting blood sugar > 120 mg/dl (Yes/No)",
    "Resting electrocardiogram results (Normal, ST-T Abnormality, LV Hypertrophy)",
    "Maximum heart rate achieved during exercise",
    "Excercise-induced angina (Yes/No)",
    "ST depression induced by exercise relative to rest",
    "Slope of the peak ST segment",
    "Presence of heart disease (Yes/No)"
    )
  )

variable_dictionary %>% 
  kable(caption = "Variable Dictionary for Heart Dataset") %>% 
  kable_styling(bootstrap_options = c("striped", "bordered"), full_width = FALSE)


```

** Clinical Background for Key Heart Variables

Several variables in this dataset come from electrocardiograms (ECG) and exercise stress tests. These exercises help determine how the heart responds to physical activity and whether blood flow to the heart is reduced. 

ST in the dataset refers to the ST segment.
- ST Segment/ ST-T Wave: The ST segment is part of the ECG waveform. Abnormal changes in this segment can indicate reduced blood flow to the heart (ischemia) or other electrical disturbances. 

- LV Hypertrophy (Left Ventricular Hypertrophy): occurs ventricle- the main pumping chamber of the heart- becomes thickened. Often resulting from long standing high blood pressure or other conditions that forces the heart to work harder. LV Hypertrophy is important because it increases the risk heart failure, and other cardiac events. 

- Oldpeak: This measure how much the ST segment drops during the exercise compared to rest. Higher values often indicate that the heart is not receiving enough oxygen.

- ST Slope: This describes the direction of the ST segment during peak exercise (upsloping, flat, downsloping). A flat or downsloping line is commonly associated with heart disease.

- Exercise Angina: Indicates whether the patient experienced chest pain during exercise, which can be a sign of reduced oxygen supply to the heart. 

Final clean dataset (condensed view)
```{r}
#-------------------------
# Sample of Cleaned Data
#-------------------------

heart %>% 
  slice_head(n = 10) %>% 
  kable(caption = "Sample of Cleaned Heart Dataset") %>% 
  kable_styling(bootstrap_options = c("striped", "bordered"), full_width = FALSE)

```

```{r outliers}
#---------------------
# Check for Outliers
#---------------------
heart %>% 
  select(age, resting_bp, cholesterol, max_hr, oldpeak) %>% 
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>% 
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot(fill = "cadetblue", alpha = 0.7) +
  labs(
    title = "Boxplots for Detecting Outliers in Key Numeric Variables",
    x = "Variable",
    y = "Value"
  ) +
  theme_minimal()


```
To identify potential outliers I created boxplots for the main numeric variables in the dataset. Variables including, age, cholesterol, max heart rate, oldpeak, resting blood pressure... In thos data we will not remove the outliers because htye reporent real cases weher the ...                                                                              
**EDA
```{r sex_count}
#-----------------------------------
# Count Bar Chart (Male vs Female)
#-----------------------------------
heart %>% 
  ggplot(aes(x = sex)) +
  geom_bar(fill = "darkgray") +
  labs(title = "Count of Male vs Female Patients",
       x = "Sex",
       y = "Count") +
  theme_minimal()
```
The bar chart shows the count of male vs female patients. This dataset contains shows significantly more males than females patients. This imbalance is important showing that raw counts alone can be misleading, hinting at sample bias, oversampling ... 

```{r heart_disease_by_sex}
#--------------------------------------------
# Heart Disease Proportion (Male vs Female)
#--------------------------------------------
heart %>% 
  ggplot(aes(x = sex, fill = heart_disease)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("No" = "#619CFF", "Yes" = "#d62728")) +
  labs(title = "Proportion of Heart Disease by Sex",
       x = "Sex",
       y = "Percentage") +
  theme_minimal()
```
This proportion bar chart shows the prevalence of heart disease in males vs females. I used a proportion chart due to the raw count imbalance- a proportion bar chart is best to showcase a direct comparison, normalizing each group to a 100%, regardless of actual count. 
The proportion chart shows that about 63%  of males have heart disease in comparison to only 26% of females that do. This correlates with real world known heart disease patterns.

** Disease vs. No Disease
* What symptoms are typically associated with heart disease? What symptoms distinguish heart disease from no heart disease?

```{r chest_pain_type_distribution}
#--------------------------------------------
# Chest Pain Type Distribution (Disease vs. No Disease)
#--------------------------------------------
heart %>% 
  ggplot(aes(x = chest_pain_type, fill = heart_disease)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Chest Pain Type by Heart Disease Status",
       x = "Chest Pain Type",
       y = "Percentage",
       fill = "Heart Disease") +
  theme_minimal()
```
We can conclude from this chart that asymptomatic chest pain type is most prevalent among individuals with heart disease, followed by typical angina, non-anginal pain, and atypical angina... this establishes chest pain type as a meaningful symptom to explore by gender.
#explain these chest pain types, in the variable dictionary sections and add it to the dict, 
# arrange it in descending order
# change color (all same?)
#add elbratory summary under the graph to remind audience of what the symptoms mean

```{r oldpeak_distribution}
#--------------------------------------------
# Oldpeak Distribution (Disease vs. No Disease)
#--------------------------------------------
heart %>% 
  ggplot(aes(x = heart_disease, y = oldpeak, fill = heart_disease)) +
  geom_boxplot() +
  
  labs(title = "Oldpeak by Heart Disease Status",
       x = "Heart Disease",
       y = "Oldpeak",
       fill = "Heart Disease") +
  theme_minimal()
```
#need to switch it out to a more informative graph (raincloud plot, bee plot) (library load), and make it more decsriptive and easier to read and understand.








** Do these symptoms differ between men and women?

```{r boxplot_correlations}


# What seems to be the highest correlation among these factors with heart disease?

# ST slope, exercise level, & oldpeak have the highest correlation. ST slope is the segment on an ECG that shows when the heart rate drops below the baseline while under physical stress, with the slope telling us the direction of that drop (depression). The oldpeak measures the drop in a numerical value. 
library(ggthemes)

# I just discovered ggthemes, gonna play with this a bit

ggplot(heart, aes(x = factor(heart_disease), y = oldpeak, fill = factor(heart_disease))) +
  geom_boxplot(alpha = 0.5) +
  labs(title = "oldpeak Distribution with Heart Disease",
       x = "Heart ") +
  theme_minimal()

```

```{r}
ggplot(heart, aes(x = heart_disease, y = oldpeak)) +
  geom_boxplot(alpha = 0.5) +
  geom_dotplot(binaxis = 'y', stackdir = "center", alpha = 0.3, binwidth = 0.2)
# Hate the way this looks. will fix later.

```





























