---
title: "Heart Prediction Model"
author: "Farrah Johnson and Mariam Ikram"
date: "2026-04-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

install.packages("knitr")
install.packages("kableExtra")
install.packages("skimr")
```
Intro: 
Heart disease is one of the leading causes of death worldwide, and the number 1 cause of death in the United States, about ... Americans from all ... (should i go into stats?) (citation) Understanding the its risk factors in pivotal for medical professionals and the general public ...

Data Exploration: (what question we're answering and exploring, background on the data: where we got it from, the link, the author, the reasoning for the data being made available, important variables and their meanings)

Packages: 
tidyverse: data wrangling (dplyr), plotting (ggplot2) and general data manipulation 
ggplot2
dyplr
janitor #cleans columns names and simple tabulations

```{r, message=FALSE, warning=FALSE, include=FALSE}

# tidyverse: data wrangling (dplyr), plotting (ggplot2), and general data manipulation (forcats)
library(tidyverse)

# janitor: cleans columns names and simple tabulations
library(janitor)

# knitr : Create nice tables in the report
library(knitr)

# kableExtra: Enhancing table appearance (border, alignment, etc.)
library(kableExtra)

library(skimr)


```

```{r import}
# Import the CSV and clean column names for easier coding
heart_raw <- read.csv("heart.csv")

# Inspect raw structure
glimpse(heart)
```
We first read in the CSV file and use glimpse() to inspect the raw imported data


```{r clean_up}

# Data Clean up

heart <- heart_raw %>% 
  clean_names() # converts column names to snake_case, lowercase, no spaces

# Map every clinical abbreviations to full text

# Sex (sex)
heart <- heart %>% 
  mutate(sex = fct_recode(factor(sex),
                          "Female" = "F",
                          "Male" = "M"
                          ))
# Check to ensure accuracy in mutation
levels(heart$sex)

# Chest Pain Type (chest_pain_type)
heart <- heart %>% 
  mutate(chest_pain_type = fct_recode(factor(chest_pain_type),
                                      "Atypical Angina" = "ATA", 
                                      "Asymptomatic" = "ASY",
                                      "Non-Anginal Pain" = "NAP", 
                                      "Typical Angina" = "TA"
                                      ))
# Check
levels(heart$chest_pain_type)

# Fasting Blood Sugar (fasting_bs), Exercise Angina (exercise_angina) Heart Disease (heart_disease)
# loop across to change names in multiple columns with one chunk
heart <- heart %>% 
  mutate(across(c(fasting_bs, exercise_angina, heart_disease), ~ {
    # Turn the current column (.x) into a factor
    fct_col <- factor(.x)
    
    # Re-code that factor
    fct_recode(fct_col,
               "Yes" = "1", "Yes" = "Y",
               "No" = "0", "No" = "N"
               )
  }))
  
# Check
levels(heart$fasting_bs)

levels(heart$exercise_angina)

levels(heart$heart_disease)

#Inspect structure after cleaning
glimpse(heart)
```
 We use `clean_names()` to standardize the column names. This makes it easier to reference variables in the rest of the analysis (for example, `chest_pain_type` instead of `ChestPainType`).  

```{r missing}

# count missing values in each column

missing_vals <- colSums(is.na(heart))
missing_vals

```
In this step we check to see if any columns have missing values. If column counts are zero, then the data has no missing value, otherwise missing values will affect how we summarize and visualize our data set columns i.e filtering out missing values or replace them with NA.

```{r duplicates}

# checking for duplicates

duplicates <- heart %>% duplicated()

duplicates_table <- duplicates %>% table()

duplicates_table


```
Here we check for duplicates in the rows of the data set using duplicated(). Duplicates can occur when data is collected from multiple sources or an error in entry, if duplicates were found I would remove them to ensure the integrity of the data. (do I talka about what I would do otehrwise? look at sample reports). Since no suplicates were found, no further action is required. 

```{r, include = FALSE}

# takes you to the full database without including it in the knit

view(heart)
```

```{r}
names(heart)
range(heart$age)
```




```{r clean_comparisons}

# Pulls out necessary data and changes the heart disease to a string to make it more decipherable 
# The 0 and 1 values in heart_disease instead of yes and no was driving me crazy. I cleaned it up and stored it in a variable.

heart_clean <- heart %>%
  mutate(heart_disease = factor(heart_disease,
                               levels = c(0, 1),
                               labels = c("No", "Yes")))

set.seed(123)
random_sample <- heart_clean %>%
  sample_n(20)
#I set a seed here so the data we're pulling is more random, as head(20) will only pull the first 20. 


random_sample %>%
  select(age, sex, heart_disease, exercise_angina, st_slope , oldpeak) %>%

  arrange(age) %>%
  head(20)

# gives an average for necessary data for a better comparison


heart_clean %>%
  select(age, sex, heart_disease, exercise_angina, st_slope , oldpeak) %>%
  
  group_by(heart_disease) %>%
  summarise("Average age" = mean(age),
            "Old Peak" = mean(oldpeak))


```

Final clean dataset (condensed view)
```{r}
# show a small sample of cleaned data
heart_clean %>% 
  slice_head(n = 10) %>% 
  kable(caption = "Sample of Cleaned Heart Dataset") %>% 
  kable_styling(full_width = FALSE)

# summary of variables of interest
#skim(heart_clean)
```


```{r boxplot_correlations}


# What seems to be the highest correlation among these factors with heart disease?

# ST slope, exercise level, & oldpeak have the highest correlation. ST slope is the segment on an ECG that shows when the heart rate drops below the baseline while under physical stress, with the slope telling us the direction of that drop (depression). The oldpeak measures the drop in a numerical value. 
library(ggthemes)

# I just discovered ggthemes, gonna play with this a bit

ggplot(heart_clean, aes(x = factor(heart_disease), y = oldpeak, fill = factor(heart_disease))) +
  geom_boxplot(alpha = 0.5) +
  labs(title = "oldpeak Distribution with Heart Disease",
       x = "Heart ") +
  theme_minimal()

```

```{r}
ggplot(heart_clean, aes(x = heart_disease, y = oldpeak)) +
  geom_boxplot(alpha = 0.5) +
  geom_dotplot(binaxis = 'y', stackdir = "center", alpha = 0.3, binwidth = 0.2)
# Hate the way this looks. will fix later.

```


```{r}
ggplot(heart_clean, aes(x = heart_disease, y = oldpeak)) +
  geom_boxplot(alpha = 0.5) +
  geom_dotplot(binaxis = 'y', stackdir = "center", alpha = 0.3, binwidth = 0.2)
# Hate the way this looks. will fix later.

```


























